---
title: 'Reproducible Research: Peer Assessment 1'
keep_md: yes
output: 
  html_document: 
    keep_md: yes
    self_contained: no
---
## Initialization
Start by setting up the parameters for external figures.
```{r}
library(knitr)
opts_chunk$set(dev="png", fig.path="figure/")
```
## Loading and preprocessing the data
Unzip and load the data from a .csv file.

```{r}
unzip("activity.zip")
activity.data <- read.csv("activity.csv")
```

Check a short summary of what have been read.
```{r}
str(activity.data)
```


The date column will be transformed to Date object so that it can be treated correctly in plots.
```{r}
activity.data$date <- as.Date(activity.data$date, format = "%Y-%m-%d")
```



## What is mean total number of steps taken per day?
Use *tapply* to calculate the number of steps taken each day.
```{r}
activity.data.dailysums <- tapply(activity.data$steps, activity.data$date, sum, na.rm = TRUE)
head(activity.data.dailysums)
```

Plot a histogram with ggplot2 showing the distribution of daily step count.
```{r}
library(ggplot2)
qplot(activity.data.dailysums, binwidth=1000, xlab="Daily step count")

```


Calculate the mean and median using the *summary* function.
```{r}
summary(activity.data.dailysums)

```




## What is the average daily activity pattern?

Use *ddply* function of the *plyr* library to calculate the mean and median values of each interval across all days: split by interval, apply mean & median, and combine to a data frame.

```{r}
library(plyr)
activity.data.daily.activity <- ddply(activity.data, c("interval"), summarize,
                                      mean = round(mean(steps, na.rm=TRUE), 2),
                                      median = round(median(steps, na.rm=TRUE), 2))
```

Plot the daily activity pattern.

```{r}
ggplot(activity.data.daily.activity, aes(x = activity.data.daily.activity$interval,
                                         y = activity.data.daily.activity$mean)) +
        geom_line() + labs(title="Mean daily activity pattern", x = "Interval", y = "Mean step count")

```

Calculate the maximum activity interval.

```{r}
activity.data.daily.activity[which.max(activity.data.daily.activity$mean), "interval"]

```


## Imputing missing values
Calculate the total number of missing values with *complete.cases* function that returns a binary vector of all the rows that have none NA's. Substract the length of that vector from the number of rows in the original data.

```{r}
nrow(activity.data) - sum(complete.cases(activity.data))
```

Fill the missing values with the mean step count of that interval.
```{r}
activity.data.complete <- ddply(activity.data, c("interval"), function(x) {
        missing <- is.na(x$steps)
        x[missing, "steps"] = mean(x[, "steps"], na.rm=TRUE)
        x
})

```

Now calculate the daily step count with the data set that has missing values imputed.

```{r}
activity.data.complete.dailysums <- tapply(activity.data.complete$steps, activity.data.complete$date, sum, na.rm = TRUE)
head(activity.data.complete.dailysums)
```
Plot a histogram of the data.
```{r}
qplot(activity.data.complete.dailysums, binwidth=1000, xlab="Daily step count")
```
The mean and median step counts can be viewed with the *summary* function.
```{r}
summary(activity.data.complete.dailysums)
```
The mean value has seen an increase compared to the original data, it is now the same as the median. The distribution now has zero skewness, which can also be seen from the histogram.


## Are there differences in activity patterns between weekdays and weekends?
Construct a new column (*$part.of.the.week*) based on lookup-table. The date variable can be transformed to literal weekday with the *weekdays* function. The literal weekday will be used as an index pointer to the lookup-table.
```{r}
part.of.the.week.lookup <- cbind(
        c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"),
        c("Weekday","Weekday","Weekday","Weekday","Weekday","Weekend","Weekend"))

activity.data.complete$part.of.the.week <- part.of.the.week.lookup[
        match(weekdays(activity.data.complete$date), part.of.the.week.lookup[,1])
        , 2]
head(activity.data.complete)
```
Now calculate the mean step count of each interval across all days with the complete data set.
```{r}
activity.data.complete.mean.intervals <- ddply(activity.data.complete, c( "interval", "part.of.the.week"),
                                               summarize,
                                               meansteps = mean(steps, na.rm = TRUE))
```
Using ggplot's facets function, it is now easy to separate the data with *part.of.the.week* variable.
```{r}
ggplot(activity.data.complete.mean.intervals) + 
        geom_line(aes(y = meansteps, x = interval), stat="identity") +
        labs(title="Mean daily activity pattern, weekdays vs. weekend", 
             x = "Interval", 
             y = "Mean step count") +
        facet_grid(facets = part.of.the.week ~ ., scales="free")
```
